---
title: "Appendix"
author: "Samuel Thau"
date: "9/26/2020"
output:
  rmarkdown::html_document:
    theme: cerulean
    
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../Posts/") })    
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# import libraries
library(tidyverse)
library(ggplot2)
library(usmap)
library(janitor)
library(usmap)
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 250px;
  overflow-y: auto;
  background-color: inherit;
}
```

## Overview
The appendix is meant to serve as a place to include information that is not suitable for the blog, like large amounts of code output, technical details, and any other information. Each section will link to the corresponding blog post. Not every blog post will have information in the appendix. 

## [3. Polls](polls.html)
```{r pollsbase, echo=FALSE,  warning = FALSE, message = FALSE }
# read in all data
nat_popvote_df <- read_csv("../Data/popvote_1948-2016.csv")
nat_poll_df    <- read_csv("../Data/pollavg_1968-2016.csv")

state_popvote_df <- read_csv("../Data/popvote_bystate_1948-2016.csv")
state_poll_df <- read_csv("../Data/pollavg_bystate_1968-2016.csv")

nat_econ <- read_csv("../Data/econ.csv")
state_rdi <- read_csv("../Data/state_rdi.csv")

#build state df with econ, polls, and 
weeks <- 5
poll_df <- state_poll_df %>% filter(weeks_left >= weeks, weeks_left<25) %>% 
  select(year, state, party, candidate_name, avg_poll) %>% 
  group_by(year, state, candidate_name, party) %>%
  summarize(avg_poll = mean(avg_poll))

#build economic df at state level from rdi and national gdp
econ_df <- state_rdi[,1:2]
colnames(econ_df) <- c("state","rdi_q2")
econ_df$rdi_q2 <- as.numeric(econ_df$rdi_q2)

for(i in seq(from = 18, to = 290, by = 16)){
  temp <- state_rdi[,c(1,i)]
  colnames(temp) <- c("state","rdi_q2")
  temp$rdi_q2 <- as.numeric(temp$rdi_q2)
  econ_df <- rbind(econ_df, temp)
}

#add years and national gdp growth
year_list <- seq(from = 1948, to = 2020, by=4)
year_append <- rep(year_list[1],51)
nat_econ_sub <- nat_econ %>% filter(year %in% year_list, quarter == 2) %>% 
  select(year, GDP_growth_qt)
gdp_append <- as.numeric(rep(nat_econ_sub[1,2],51))

for(i in seq(from = 2, to=length(year_list), by =1)){
  new_year = rep(year_list[i], 51)
  new_gdp = as.numeric(rep(nat_econ_sub[i,2], 51))
  year_append = append(year_append, new_year)
  gdp_append = append(gdp_append, new_gdp)
}

#append year and national gdp to the econ df
econ_df$year <- year_append
econ_df$gdp <- gdp_append


## Merge datasets and clean
inc_data <- nat_popvote_df %>% select("year","party","winner", "incumbent","incumbent_party")

reg_df <- poll_df %>% left_join(econ_df, by=c("state"="state","year"="year")) %>% 
  left_join(state_popvote_df, by= c("state"="state", "year"="year")) %>%
  left_join(inc_data, 
            by  = c("year"="year","party" = "party" ))
  
reg_df$period <- ifelse(reg_df$year %in% 1972:1984, "1972-1984",
                               ifelse(reg_df$year %in% 1988:1996, "1992-2000",
                                      ifelse(reg_df$year %in% 2000:2012, "2000-2012", 
                                             "2016-2020")))

reg_df <- reg_df %>% mutate("d_pv"=100*D/total) %>% mutate("r_pv"=100*R/total)
reg_df <- reg_df %>% 
  mutate(inc_party = case_when((party=="democrat" & incumbent_party==TRUE) ~ "democrat", 
                                 (party=="democrat" & incumbent_party==FALSE) ~ "republican",
                                 (party=="republican" & incumbent_party==TRUE) ~ "republican",
                                 (party=="republican" & incumbent_party==FALSE) ~ "democrat")) %>%
  mutate(chl_party = case_when((party=="democrat" & incumbent_party==FALSE) ~ "democrat", 
                               (party=="democrat" & incumbent_party==TRUE) ~ "republican",
                               (party=="republican" & incumbent_party==FALSE) ~ "republican",
                               (party=="republican" & incumbent_party==TRUE) ~ "democrat"))

reg_df <- reg_df %>% 
  mutate(inc_pv = case_when((inc_party=="republican") ~ r_pv, 
                              (inc_party=="democrat") ~ d_pv)) %>% 
  mutate(chl_pv = case_when(chl_party == "democrat" ~ d_pv, 
                            chl_party == "republican" ~ r_pv))

reg_df <- na.omit(reg_df)

##fundamentals model
inc_fund_lm <- lm(inc_pv ~ rdi_q2 + gdp + period + state, data = reg_df %>% filter(year < 2020, incumbent_party == TRUE))
#summary(inc_fund_lm)

chl_fund_lm <- lm(chl_pv ~ rdi_q2 + gdp + period + state, data = reg_df %>% filter(year < 2020, incumbent_party == FALSE))
#summary(chl_fund_lm)

## polls only model
inc_polls_lm <- lm(inc_pv ~ avg_poll, data = reg_df %>% filter(year < 2020, incumbent_party == TRUE))
chl_polls_lm <- lm(chl_pv ~ avg_poll, data = reg_df %>% filter(year < 2020, incumbent_party == FALSE))

## polls and fundamentals model
inc_plus_lm <- lm(inc_pv ~ rdi_q2 + gdp + period + state + avg_poll, data = reg_df %>% filter(year < 2020, incumbent_party == TRUE))
#summary(inc_plus_lm)

chl_plus_lm <- lm(chl_pv ~ rdi_q2 + gdp + period + state + avg_poll, data = reg_df %>% filter(year < 2020, incumbent_party == FALSE))
#summary(chl_plus_lm)

```

During the blog post for polls, I built two linear models: a so called fundamentals model that is based on the economy, and the polls plus model that adds in the impact of vote shares. While the most important information was discussed in the blog post, there is some value in seeing the full regression output which is provided below. These are two sided models, meaning that there is a seperate model for candidates from the incumbent party, and from the challenging party.First, the updated fundamentals model for the incumbent:
```{r polls_fund_inc_lm, echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(inc_fund_lm)
```
Then, for the challenger:
```{r polls_fund_chl_lm, echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(chl_fund_lm)
```
I also built the polls plus model, that uses the fundamentals and a polling average as predictors in a linear model. For the incumbent, the regression results are as follows:
```{r polls_plus_inc_lm, echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(inc_plus_lm)
```
Then, we can also look at the results for the challengers:
```{r polls_plus_chl_lm, echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(chl_plus_lm)
```


## [5. Probabilistic Models](polls1.html)
```{r , echo=FALSE,  warning = FALSE, message = FALSE }
rm(list = ls())
set.seed(1234)

# read in all data
nat_popvote_df <- read_csv("../Data/popvote_1948-2016.csv")
nat_poll_df    <- read_csv("../Data/pollavg_1968-2016.csv")

state_popvote_df <- read_csv("../Data/popvote_bystate_1948-2016.csv")
state_poll_df <- read_csv("../Data/pollavg_bystate_1968-2016.csv")

nat_econ <- read_csv("../Data/econ.csv")
state_rdi <- read_csv("../Data/state_rdi.csv")

vep_df <- read_csv("../Data/vep_1980-2016.csv")

# clean data
#build state df with econ, polls, and 
weeks <- 4
poll_df <- state_poll_df %>% filter(weeks_left >= weeks, weeks_left<25) %>% 
  select(year, state, party, candidate_name, avg_poll) %>% 
  group_by(year, state, candidate_name, party) %>%
  summarize(avg_poll = mean(avg_poll))

#build economic df at state level from rdi and national gdp
econ_df <- state_rdi[,1:2]
colnames(econ_df) <- c("state","rdi_q2")
econ_df$rdi_q2 <- as.numeric(econ_df$rdi_q2)

for(i in seq(from = 18, to = 290, by = 16)){
  temp <- state_rdi[,c(1,i)]
  colnames(temp) <- c("state","rdi_q2")
  temp$rdi_q2 <- as.numeric(temp$rdi_q2)
  econ_df <- rbind(econ_df, temp)
}

#add years and national gdp growth
year_list <- seq(from = 1948, to = 2020, by=4)
year_append <- rep(year_list[1],51)
nat_econ_sub <- nat_econ %>% filter(year %in% year_list, quarter == 2) %>% 
  select(year, GDP_growth_qt)
gdp_append <- as.numeric(rep(nat_econ_sub[1,2],51))

for(i in seq(from = 2, to=length(year_list), by =1)){
  new_year = rep(year_list[i], 51)
  new_gdp = as.numeric(rep(nat_econ_sub[i,2], 51))
  year_append = append(year_append, new_year)
  gdp_append = append(gdp_append, new_gdp)
}

#append year and national gdp to the econ df
econ_df$year <- year_append
econ_df$gdp <- gdp_append


## Merge datasets and clean
inc_data <- nat_popvote_df %>% select("year","party","winner", "incumbent","incumbent_party")

reg_df <- poll_df %>% left_join(econ_df, by=c("state"="state","year"="year")) %>% 
  left_join(state_popvote_df, by= c("state"="state", "year"="year")) %>%
  left_join(inc_data, 
            by  = c("year"="year","party" = "party" ))

reg_df$period <- ifelse(reg_df$year %in% 1952:1972, "1952-1976",
                        ifelse(reg_df$year %in% 1972:1992, "1972-1992", "1996-2020"))

reg_df <- reg_df %>% mutate("d_pv"=100*D/total) %>% mutate("r_pv"=100*R/total)
reg_df <- reg_df %>% 
  mutate(inc_party = case_when((party=="democrat" & incumbent_party==TRUE) ~ "democrat", 
                               (party=="democrat" & incumbent_party==FALSE) ~ "republican",
                               (party=="republican" & incumbent_party==TRUE) ~ "republican",
                               (party=="republican" & incumbent_party==FALSE) ~ "democrat")) %>%
  mutate(chl_party = case_when((party=="democrat" & incumbent_party==FALSE) ~ "democrat", 
                               (party=="democrat" & incumbent_party==TRUE) ~ "republican",
                               (party=="republican" & incumbent_party==FALSE) ~ "republican",
                               (party=="republican" & incumbent_party==TRUE) ~ "democrat"))

reg_df <- reg_df %>% 
  mutate(inc_pv = case_when((inc_party=="republican") ~ R, 
                            (inc_party=="democrat") ~ D)) %>% 
  mutate(chl_pv = case_when(chl_party == "democrat" ~ D, 
                            chl_party == "republican" ~ R))

reg_df <- reg_df %>% mutate(incumbent_win = case_when(inc_party == "republican" & r_pv > d_pv ~ 1,
                                                      inc_party == "republican" & r_pv < d_pv ~ 0, 
                                                      inc_party == "democrat" & r_pv < d_pv ~ 1,
                                                      inc_party == "democrat" & r_pv > d_pv ~ 0))

reg_df <- reg_df %>% left_join(vep_df, by = c("state" = "state", "year" = "year"))

reg_df <- reg_df %>% filter(year > 1979)

reg_df <- na.omit(reg_df)


## logit regression for probability
inc_log <- glm(cbind(inc_pv, VEP-inc_pv) ~ rdi_q2 + gdp + state + avg_poll + period, 
               data = reg_df %>% filter(year < 2020, incumbent_party == TRUE), family=binomial)

chl_log <- glm(cbind(chl_pv, VEP-chl_pv) ~ rdi_q2 + gdp + state + avg_poll + period, 
               data = reg_df %>% filter(year < 2020, incumbent_party == FALSE), family=binomial)

```
In this post, I further explored probabilistic models, focusing on a logistic regression with binomial output. I built a two sided model, looking at both incumbent and challenger outcomes, predicting the number of votes for a candidate given the size of the voter eligible population at the state level. 

The logistic regression is of the following form:

$$ (Votes | VEP)= f(\beta_0 + \beta_1 RDI + \beta_2 GDP + \beta_3 Poll + \beta_4 Period +\beta_5 State)$$
Where $f(x)=\frac{exp(x)}{1+exp(x)}$. In addition, VEP is the eligible voting popluation, RDI is the second quarter RDI in a particular state, GDP is the national second quarter GDP, poll is the average vote share up to 4 weeks before the election, period is an indicator for 1980-1992 or 1996 onward, and state is a set of dummy variables for states. We can exmaine the output for both the incumbent party:

```{r , echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(inc_log)
```

And also the challenger party.

```{r , echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(chl_log)
```

The deviances (one way of measuring the fit of a logistic regression) are extremely high, suggesting a poor fit. In addition to the problems with a [lack of variance](polls1.html), it would seem that this model may not be particularly helpful. 

## [7. Parties and Uncertainty](polarization.html)
In the post on parties and uncertainty, I used a number of logistic models based on two different types of two sided model: one based on incumbency status, and one based on parties. The results of those regressions are as follows. 

```{r , echo=FALSE,  warning = FALSE, message = FALSE }
##libraries
library(tidyverse)
library(ggplot2)
library(usmap)
library(janitor)
library(usmap)
library(data.table)
library(geofacet) ## map-shaped grid of ggplots
library(stargazer)


# read in all data
nat_popvote_df <- read_csv("../Data/popvote_1948-2016.csv")
nat_poll_df    <- read_csv("../Data/pollavg_1968-2016.csv")

state_popvote_df <- read_csv("../Data/popvote_bystate_1948-2016.csv")
state_poll_df <- read_csv("../Data/pollavg_bystate_1968-2016.csv")

nat_econ <- read_csv("../Data/econ.csv")
state_rdi <- read_csv("../Data/state_rdi.csv")

vep_df <- read_csv("../Data/vep_1980-2016.csv")

# clean data
#build state df with econ, polls, and 
weeks <- 1
poll_df <- state_poll_df %>% filter(weeks_left >= weeks, weeks_left<5) %>% 
  select(year, state, party, candidate_name, avg_poll) %>% 
  group_by(year, state, candidate_name, party) %>%
  summarize(avg_poll = mean(avg_poll))

#build economic df at state level from rdi and national gdp
econ_df <- state_rdi[,1:2]
colnames(econ_df) <- c("state","rdi_q2")
econ_df$rdi_q2 <- as.numeric(econ_df$rdi_q2)

for(i in seq(from = 18, to = 290, by = 16)){
  temp <- state_rdi[,c(1,i)]
  colnames(temp) <- c("state","rdi_q2")
  temp$rdi_q2 <- as.numeric(temp$rdi_q2)
  econ_df <- rbind(econ_df, temp)
}

#add years and national gdp growth
year_list <- seq(from = 1948, to = 2020, by=4)
year_append <- rep(year_list[1],51)
nat_econ_sub <- nat_econ %>% filter(year %in% year_list, quarter == 2) %>% 
  select(year, GDP_growth_qt)
gdp_append <- as.numeric(rep(nat_econ_sub[1,2],51))

for(i in seq(from = 2, to=length(year_list), by =1)){
  new_year = rep(year_list[i], 51)
  new_gdp = as.numeric(rep(nat_econ_sub[i,2], 51))
  year_append = append(year_append, new_year)
  gdp_append = append(gdp_append, new_gdp)
}

#append year and national gdp to the econ df
econ_df$year <- year_append
econ_df$gdp <- gdp_append


## Merge datasets and clean
pol_data <- nat_popvote_df %>% select("year","party","winner", "incumbent","incumbent_party")

reg_df <- poll_df %>% left_join(econ_df, by=c("state"="state","year"="year")) %>% 
  left_join(state_popvote_df, by= c("state"="state", "year"="year")) %>%
  left_join(pol_data, 
            by  = c("year"="year","party" = "party" ))


reg_df <- reg_df %>% 
  mutate(inc_party = case_when((party=="democrat" & incumbent_party==TRUE) ~ "democrat", 
                               (party=="democrat" & incumbent_party==FALSE) ~ "republican",
                               (party=="republican" & incumbent_party==TRUE) ~ "republican",
                               (party=="republican" & incumbent_party==FALSE) ~ "democrat")) %>%
  mutate(chl_party = case_when((party=="democrat" & incumbent_party==FALSE) ~ "democrat", 
                               (party=="democrat" & incumbent_party==TRUE) ~ "republican",
                               (party=="republican" & incumbent_party==FALSE) ~ "republican",
                               (party=="republican" & incumbent_party==TRUE) ~ "democrat"))

reg_df <- reg_df %>% 
  mutate(inc_pv = case_when((inc_party=="republican") ~ R_pv2p, 
                            (inc_party=="democrat") ~ D_pv2p)) %>% 
  mutate(chl_pv = case_when(chl_party == "democrat" ~ D_pv2p, 
                            chl_party == "republican" ~ R_pv2p))

reg_df <- reg_df %>% mutate(incumbent_win = case_when(inc_party == "republican" & R_pv2p > D_pv2p ~ 1,
                                                      inc_party == "republican" & R_pv2p < D_pv2p ~ 0, 
                                                      inc_party == "democrat" & R_pv2p < D_pv2p ~ 1,
                                                      inc_party == "democrat" & R_pv2p > D_pv2p ~ 0))

reg_df <- reg_df %>% 
  mutate(inc_votes = case_when((inc_party=="republican") ~ R, 
                            (inc_party=="democrat") ~ D)) %>% 
  mutate(chl_votes = case_when(chl_party == "democrat" ~ D, 
                            chl_party == "republican" ~ R))
reg_df <- reg_df %>% left_join(vep_df, by = c("state" = "state", "year" = "year"))

reg_df <- na.omit(reg_df)

reg_df <- reg_df %>% mutate_at(vars(VEP), funs(round(.)))

## do regressions for in sample fit
inc_log_base <- glm(cbind(inc_votes, VEP-inc_votes) ~ rdi_q2 + gdp + state + avg_poll, 
               data = reg_df %>% filter(year < 2020, incumbent_party == TRUE), family=binomial)

chl_log_base <- glm(cbind(chl_votes, VEP-chl_votes) ~ rdi_q2 + gdp + state + avg_poll, 
               data = reg_df %>% filter(year < 2020, incumbent_party == FALSE), family=binomial)

inc_log_cont <- glm(cbind(inc_votes, VEP-inc_votes) ~ rdi_q2 + gdp + state + avg_poll+ inc_party, 
                    data = reg_df %>% filter(year < 2020, incumbent_party == TRUE), family=binomial)

chl_log_cont <- glm(cbind(chl_votes, VEP-chl_votes) ~ rdi_q2 + gdp + state + avg_poll + inc_party, 
                    data = reg_df %>% filter(year < 2020, incumbent_party == FALSE), family=binomial)

## summaries of models
# summary(inc_log_base)
# summary(chl_log_base)
# summary(inc_log_cont)
# summary(chl_log_cont)


##
## do regressions for in sample fit
dem_log_base <- glm(cbind(D, VEP-D) ~ rdi_q2 + gdp + state + avg_poll, 
                    data = reg_df %>% filter(year < 2020, party == "democrat"), family=binomial)

rep_log_base <- glm(cbind(R, VEP-R) ~ rdi_q2 + gdp + state + avg_poll, 
                    data = reg_df %>% filter(year < 2020, party == "republican"), family=binomial)

dem_log_cont <- glm(cbind(D, VEP-D) ~ rdi_q2 + gdp + state + avg_poll+ incumbent_party, 
                    data = reg_df %>% filter(year < 2020, party == "democrat"), family=binomial)

rep_log_cont <- glm(cbind(R, VEP-R) ~ rdi_q2 + gdp + state + avg_poll + incumbent_party, 
                    data = reg_df %>% filter(year < 2020, party == "republican"), family=binomial)

## summaries of models
# summary(dem_log_base)
# summary(rep_log_base)
# summary(dem_log_cont)
# summary(rep_log_cont)


```

First look at the model for the incumbent party. 

```{r , echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(inc_log_cont)
```

Then the challenger party. 
```{r , echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(chl_log_cont)
```

We can then switch over to the party based model. First look at the Democrat side. 

```{r , echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(dem_log_cont)
```

Then, finally, we can look at the Republcan side. 

```{r prob1_chl, echo=FALSE,  warning = FALSE, message = FALSE, class.output="scroll-100" }
 summary(rep_log_cont)
```



